\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-07 forwards-and-backwards.tex
Update on Overleaf.

% % % % % % % 
\subsubsection*{Mean measure}

Consider first the mean measure.
The birth rate at location $x$ when the state of the population is $\eta$
is a function of the location and the measure $\eta$ through a specific form:
$\gamma(x, \smooth{\gamma} \eta(x))$;
however, for the moment we will write $\gamma(x, \eta)$ instead
to make the exposition clearer
(and similarly for establishment probability $r(x, \eta)$ and death rate $\mu(x, \eta)$).
Recall that in our parameterization only $\mu$ and the dispersal kernel $q$ depend on $\theta$.
For a suitable (smooth, compactly supported) test function $f$, we find
\begin{equation} \label{mean measure}
    \begin{split}
    \Pgen^\theta \langle f, \eta \rangle
    &=
    \lim_{\delta t\downarrow 0} \frac{1}{\delta t}
    \IE\left[ \left.
        \langle f, \eta_{\delta t} \rangle
        -
        \langle f, \eta\rangle
        \right| \eta_0 = \eta
    \right]
    \\
    &=
    \theta \int
        \int f(z) r(z,\eta) q_\theta(x,dz)
    \gamma(x, \eta) \eta(dx)
    -
    \theta \int f(x)\mu_\theta(x, \eta)
    \eta(dx).
    \end{split}
\end{equation}
The first term is the increment in $\langle f,\eta\rangle$
resulting from a birth event (recalling that
we don't kill the parent) integrated against the rate of such events,
and the second reflects death events.
In both terms,
the rate of events has a factor of $N$ (because events happen at a rate 
proportional to the number of individuals,
whereas $\eta$ has mass $1/N$ for each individual)
which is offset by the fact that  
the birth or loss of a single 
individual at the point $y$, say, changes $\langle f,\eta\rangle$
by $f(y)/N$.

We use the fact that $\int q_\theta(x,dz)=1$ to rewrite~(\ref{mean measure})
as 
\begin{equation}
\label{eqn:rewritten mean measure}
\begin{split}
    \int\left(
        \int \theta \left( f(z) r(z,\eta)- f(x) r(x,\eta) \right) q_\theta(x,dz)
    \right)
    \gamma(x,\eta)
    \eta(dx)
    \\
    + \int \int f(x) \theta \Big(
        r(x,\eta) \gamma(x,\eta)
        - \mu_\theta(x,\eta)
    \Big) \eta(dx).
\end{split}
\end{equation}
We have defined $\mu_\theta$ so that the second term is simple:
\begin{align*}
    \theta \Big( r(x,\eta) \gamma(x,\eta) - \mu_\theta(x,\eta) \Big)
    = F(x, \eta) .
\end{align*}
This definition --
is really a `near-critical' assumption that birth and death rates
differ by something of order $1/\theta$.
Furthermore, since we've taken $q_\theta$ to be the density of a Normal
with mean and variance proportional to $1/\theta$,
\begin{align} \label{eqn:near_critical}
    \int \theta \Big(
        r(z,\eta) f(z)
        -
        r(x,\eta) f(x)
    \Big) q_\theta(x,dz) 
    \qquad \stackrel{\theta\to\infty}{\longrightarrow} \qquad  
    \DG \big(r(\cdot,\Xi)f(\cdot)\big)(x),
\end{align}
where recall if dispersal is simply standard multivariate Gaussian
with mean zero and covariance $\sigma^2 I / \theta$,
then $\DG = \sigma^2 \Delta$, where $\Delta$ denotes the Laplacian.
Note that there is a slight mismatch between the locations at which we are 
measuring birth and death rates, because in~\eqref{eqn:near_critical}
the establishment probability is 
measured at $x$ rather than at the point where juveniles fall. 
\comment{(IS THIS BIT STILL TRUE:)}
This condition also differs from the assumption that
we use in~\S\ref{scalings with levels} to guarantee convergence of the 
processes with levels in our lookdown 
constructions, see~(\ref{lookdown blambda}). 
Under mild
regularity conditions on $r$ (which we certainly need to have sensible 
limiting models) the conditions are equivalent.

Equation~\eqref{eqn:rewritten mean measure}) then converges to
\begin{equation} \label{limit of mean measure equation}
\int \gamma(x,\eta)
\DG \big(f(\cdot)r(\cdot,\eta)\big)(x)
\eta(dx)
+
\int f(x)
F(x,\eta)
\eta(dx) .
\end{equation}
More generally, we expect everything to work
if the remaining parameters ($r$, $\gamma$, and $F$)
depended on $\theta$, but converged as $\theta \to \infty$.

% % % % % % % 
\subsubsection*{Quadratic variation}

% Note: this bit in firstversionplus.tex has things written out in detail for
% the Poisson offspring number case.

We now look at the second order term.
An individual at location $x$ gives birth 
to a surviving offspring at $y$ at rate
$$
\gamma(x,\eta) r(y,\eta) q_{\theta}(x, dy) ,
$$
and since this increments $\langle f, \eta \rangle$ by $f(y) / N$,
the contribution to the quadratic variation from birth events,
which occur at rate $\theta$ per individual 
(so, rate $N\theta |\eta|$ overall), is
$$
\langle
    N \theta \gamma(x,\eta)
    \int \frac{1}{N^2} f^2(y) r(y,\eta)
    q_\theta(x,dy) 
    , \eta(dx)
\rangle .
$$
Similarly, the increment in $\langle f, \eta\rangle$ resulting from 
the death of an individual at $x$ is $f(x)/N$, and so combining with the 
above, the second order term in the generator takes the form
$$
F''(\langle f,\eta\rangle)
\frac{1}{2} N \theta
\left\{
    \langle
        \gamma(x,\eta)
        \int \frac{1}{N^2}f^2(y)r(y,\eta)q_\theta(x,dy) 
    , \eta(dx)\rangle
    +
    \langle
        \mu_\theta(x,\eta)\frac{1}{N^2}f^2(x) 
    ,\eta(dx)\rangle
\right\} .
$$
The term $\varepsilon_{\theta,N}(f, F, \eta)$ will be 
$\bigO(\theta/N^2)$.

% % % % % % % %
\subsubsection*{Limit I: deterministic case}

\comment{Need to decide what to call the limit here (and above, I guess).}

By choosing $\theta/N\rightarrow 0$, the second order term in the generator 
will vanish and we expect a deterministic limit. 
Let us suppose that the limit has a smooth density with respect to
Lebesgue measure, and abuse notation by using $\Xi_t(x)$
to denote the corresponding density, then integrating the first term in~(\ref{limit of mean measure equation})
by parts, we find that in the limit as $N\to\infty$,
\begin{equation*}
\frac{d}{dt}\int f(x)\Xi_t(x)dx
=
\int f(x)\left\{r(x,\Xi_t)\DG\big(\gamma(\cdot,\Xi_t)
\Xi_t(\cdot)\big)(x)
+
F(x,\Xi_t) \Xi_t(x)
\right\} dx.
\end{equation*}
In other words, we can expect the limit to be a weak solution to the 
deterministic equation
\begin{equation}
\label{deterministic limit}
\frac{d\Xi}{dt}
=r(x,\Xi)\DG\big(\gamma(\cdot,\Xi)
\Xi(\cdot)\big)(x)+F(x,\Xi)
\Xi(x).
\end{equation}

@@

If $N=\delta\theta$, for some positive constant $\delta$,
the second order term in the generator remains order
one as $\theta\to\infty$, and  
we expect a `generalised superprocess' limit.
Assuming some regularity of the coefficients (in particular of $r$), 
the second order term in the generator will take the form
$$
\frac{1}{\delta}F''(\langle f,\Xi\rangle)
\langle f^2(x), \gamma(x,\Xi)
r(x,\Xi)\Xi(dx)\rangle.
$$
That is
$$\langle f(x), \Xi_t(dx)\rangle-\langle f(x), \Xi_0(dx)\rangle
-\int_0^t\langle\gamma(x,\Xi_s)\DG\big(r(\cdot,\Xi_s)f(\cdot)\big)(x),
\Xi_s(dx)\rangle ds$$
is a martingale with quadratic variation
$$\frac{2}{\delta}\int_0^t\langle f^2(x) 
\gamma(x,\Xi_s)r(x,\Xi_s),\Xi_s(dx)\rangle ds.$$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Technical Lemmas}
\subsection{Proof to Lemma \ref{lem: gamma - r radius comparison}}

>>>>

\subsubsection*{Mean measure}

Consider first the mean measure. For a suitable 
(smooth, compactly supported) test function $f$, we find
\begin{multline}
\label{mean measure}
{\cal L}^{(\theta, N)}\langle f,\eta\rangle=
\lim_{\delta t\downarrow 0}\frac{1}{\delta t}
\IE\left[\left.\langle f,\eta_{\delta t}\rangle -
\langle f, \eta\rangle \right|\eta_0=\eta
\right]\\
=\theta\int\int f(z)r(z,\eta)q_\theta(x,dz)
\gamma(x,\eta)
\eta^{\theta}_s(dx)-\theta\int f(x)\mu(x,\eta)
\eta^{\theta}_s(dx).
\end{multline}
The first term is the increment in $\langle f,\eta\rangle$
resulting from a birth event (recalling that
we don't kill the parent) integrated against the rate of such events,
and the second reflects death events.
In both terms, the 
rate of events has a factor of $N$ (because events happen at a rate 
proportional to the number of individuals, whereas $\eta$ is
the scaled population mass) which is offset by the fact that  
the birth or loss of a single 
individual at the point $y$, say, changes $\langle f,\eta\rangle$
by $f(y)/N$.

We use the fact that $\int q_\theta(x,dz)=1$ to rewrite~(\ref{mean measure})
as 
\begin{multline}
\label{rewritten mean measure}
\int\left(\int \theta \left(
f(z)r(z,\eta)- f(x)r(x,\eta)\right)
q_\theta(x,dz)\right)
\gamma(x,\eta)
\eta^{\theta}_s(dx)
\\
+\int\theta\int f(x)\Big(r(x,\eta)
\gamma(x,\eta)
-\mu(x,\eta)\Big)
\eta^{\theta}_s(dx).
\end{multline}
Suppose that $\eta\to\Xi$, a measure on $\IR^d$,
in the scaling limit, then we see that we can 
let $\theta\rightarrow\infty$ in~(\ref{rewritten mean measure})
under the assumptions that
\begin{eqnarray}
r(x,\eta)&\stackrel{\theta\to\infty}{\longrightarrow}&
r(x,\Xi)\\
\gamma(x,\eta)&\stackrel{\theta\to\infty}{\longrightarrow}&
\gamma(x,\Xi)\\
\int\theta \Big(r(z,\eta)f(z)-
r(x,\eta)f(x)\Big)q_\theta(x,dz) 
&\stackrel{\theta\to\infty}{\longrightarrow}& 
\Delta \big(r(\cdot,\Xi)f(\cdot)\big)(x);\\
\theta \Big(r(x,\eta)\gamma (x,\eta)
-\mu (x,\eta)\Big)
 &\stackrel{\theta\to\infty}{\longrightarrow}& 
F(x,\Xi),
\label{near critical}
\end{eqnarray}
where $\Delta$ denotes the Laplacian.
The last condition reflects the `near-critical' branching. Note that
there is a slight mismatch between the locations at which we are 
measuring birth and death rates, because in~(\ref{near critical})
the establishment probability is 
measured at $x$ rather than at the point where juveniles fall. 
This condition also differs from the assumption that
we use in~\S\ref{scalings with levels} to guarantee convergence of the 
processes with levels in our lookdown 
constructions, see~(\ref{lookdown blambda}). 
Under mild
regularity conditions on $r$ (which we certainly need to have sensible 
limiting models) the conditions are equivalent.

Equation~(\ref{rewritten mean measure}) converges to
\begin{equation}%\begin{multline}
\label{limit of mean measure equation}
\int \gamma(x,\Xi)
\Delta
\big(f(\cdot)r(\cdot,\Xi)\big)(x)
\Xi(dx)
+\int f(x)
F(x,\Xi)
\Xi(dx)
.
\end{equation}

\subsubsection*{Quadratic variation}

We now look at the second order term.
We spell out
this calculation in detail so that it is evident at which points we are 
exploiting the choice of the Poisson distribution for the number
of juvenile offspring.
In the general case one can expect somewhat messier calculations, but
the same conclusion (up to constants).

When an individual at the point $x$ gives birth, the total number of
mature offspring is Poisson with parameter
$$\gamma(x,\eta)\int 
r(y,\eta) q_{\theta}(x, dy)
=\gamma(x,\eta)p_\theta(x,\eta).$$
Offspring that survive to maturity are distributed about the parent
according to 
$$\overline{q}_\theta(x,\eta,dy):=
\frac{r(y,\eta)q_\theta(x,dy)} 
{\int r(z,\eta)q_\theta(x,dz)}. 
$$
Writing $X(x)$ for the increment to $\langle f, \eta\rangle$
due to a single mature offspring of an individual at $x$, 
and $P(x)$ for the (Poisson) number of mature
offspring in such an event, the variance of the increment to 
$\langle f, \eta\rangle$
resulting from a birth event at $x$ is $\IE[P(x)]\mathtt{var}(X(x))
+\mathtt{var}(P(x))\IE[X(x)]^2=\IE[P]\IE[X(x)^2]$. 
Since, in this notation, conditional on the value of $\eta$
immediately before the event,
$$\IE[X(x)^2]=\int
\frac{1}{N^2}f^2(y) \overline{q}_\theta(x,\eta,dy),$$
and
$$\IE[P(x)]=
\gamma (x,\eta) \int 
r(z,\eta)q_\theta(x,dz),$$ 
the contribution to the quadratic variation from birth events, which
occur at rate $\theta$ per individual 
(so $N\theta |\eta|$ overall), is
$$\langle N\theta \gamma(x,\eta)
\int \frac{1}{N^2}f^2(y)r(y,\eta)
q_\theta(x,dy) 
,\eta^{\theta}_s(dx)\rangle.$$
The increment in $\langle f, \eta\rangle$ resulting from 
the death of an individual at $x$ is $f(x)/N^2$, and so combining with the 
above, the second order term in the generator takes the form
$$F''(\langle f,\eta\rangle)
\frac{1}{2}N\theta \left\{
\langle \gamma(x,\eta)
\int \frac{1}{N^2}f^2(y)r(y,\eta)q_\theta(x,dy) 
,\eta^{\theta}_s(dx)\rangle
+\langle \mu(x,\eta)\frac{1}{N^2}f^2(x) 
,\eta^{\theta}_s(dx)\rangle\right\}.
$$
The term $\varepsilon_{\theta,N}(f, F, \eta)$ will be 
${\mathcal O}(\theta/N^2)$.

\subsubsection*{Limit I: deterministic case}

By choosing $\theta/N\rightarrow 0$, the second order term in the generator 
will vanish and we expect a deterministic limit. 
Let us suppose that the limit has a smooth density with respect to
Lebesgue measure, and abuse notation by using $\Xi_t(x)$
to denote the corresponding density, then integrating the first term in~(\ref{limit of mean measure equation})
by parts, we find that in the limit as $N\to\infty$,
\begin{equation}
\nonumber
\frac{d}{dt}\int f(x)\Xi_t(x)dx
=\int f(x)\left\{r(x,\Xi_t)\Delta\big(\gamma(\cdot,\Xi_t)
\Xi_t(\cdot)\big)(x)+F(x,\Xi_t)
\Xi_t(x)
\right\}dx.
\end{equation}
In other words, we can expect the limit to be a weak solution to the 
deterministic equation
\begin{equation}
\label{deterministic limit}
\frac{d\Xi}{dt}
=r(x,\Xi)\Delta\big(\gamma(\cdot,\Xi)
\Xi(\cdot)\big)(x)+F(x,\Xi)
\Xi(x).
\end{equation}

@@

If $N=\rho\theta$, for some constant $\rho$,
the second order term in the generator remains order
one as $\theta\to\infty$, and  
we expect a `generalised
superprocess'
limit. Assuming some regularity of the coefficients (in particular 
of $r$), 
the second order term in the generator will  
take the form
$$\frac{1}{\rho}F''(\langle f,\Xi\rangle)
\langle f^2(x), \gamma(x,\Xi)
r(x,\Xi)\Xi(dx)\rangle.
$$
That is
$$\langle f(x), \Xi_t(dx)\rangle-\langle f(x), \Xi_0(dx)\rangle
-\int_0^t\langle\gamma(x,\Xi_s)\Delta\big(r(\cdot,\Xi_s)f(\cdot)\big)(x),
\Xi_s(dx)\rangle ds$$
is a martingale with quadratic variation
$$\frac{2}{\rho}\int_0^t\langle f^2(x) 
\gamma(x,\Xi_s)r(x,\Xi_s),\Xi_s(dx)\rangle ds.$$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Technical Lemmas}
\subsection{Proof to Lemma \ref{lem: gamma - r radius comparison}}
